Mergeable Types

### 2017-02-27 paper/examples.tex
EDits.

{\bf Ropes}. A rope is a data structure that faciliates efficient
implementations of operations such as concatenation, splitting, lookup
and insertion at a random position etc., for a list of values. The
values are usually characters; their lists being strings. The data
structure is a binary tree, whose leaf nodes store substrings of a
larger string such that their left-to-right (in-order) concatenation
yeilds the larger string. The idea is to facilitate efficient lookups
and insertions at random positions in the string by storing
position indexes in the internal nodes of the tree.

@@

@@

{\bf Lists}. As an abstract data type, lists support three
operations: \C{insert x i l}, to insert an element \C{x} at position
\C{i} in the list \C{l}, \C{delete i l}, to delete the element at
\C{i}'th position, and \C{subst i x l}, to substitute the element at
the \C{i}'th position with \C{x}.  Lists of mergeable items are
mergeable if they preserve the following properties:
\begin{itemize}
  \item Elements present in both the merging lists will be present in
  the final list. No element is deleted unless it is deleted in at
  least one list.
  \item An element deleted in one list will not reappear in the merged
  list (unlike Amazon's replicated shopping cart~\cite{dynamo}).
  \item The order of elements is retained. That is, if $x$ occurs
  before $y$ in one of the lists, and if both elements are present in
  the merged list, then $x$ occurs before $y$ in the merged list.
\item If an element $y$ is substituted for an element $x$ in one list,
  then $y$ is also substituted for $x$ in the merged list (i.e., $x$
  is absent and $y$ is present), unless $x$ is deleted or substituted
  in the other list. In the former case, deletion wins, and neither
  $x$ nor $y$ occur in the merged list. In the latter case, if $x$ is
  substituted by a different element $z$ in the other list, then the
  merged list substitutes $x$ by a merge of $y$ and $z$, defined as
  per the merge semantics of the list element type.
\end{itemize}

@@

A significant advantage of mergeable types is their composability. As
demonstrated by \C{MList}, the merge semantics of a composed mergeable
type is a composition of the merge operations of its constituent types.  For
instance, a mergeable list of integers can be obtained by
instantiating \C{MList} for an \C{Int} type, where \C{Int.merge} is
the same as \C{Counter.merge} (Sec.~\ref{sec:intro}).
An alternative instantiation of \C{MList} is for a quantity type
\begin{wrapfigure}{l}{.45\textwidth}
  \centering
  \includegraphics[scale=0.4]{Figures/list-eg}
  \caption{Lists of mergeable values are mergeable. }
  \label{fig:list-eg}
\end{wrapfigure}
(\C{Qty}) that supports \C{add} and \C{subtract}
operations, but defines its \C{merge} opeation a little differently:
\begin{ocaml}
let merge lca v1 v2 = if v1 = lca then v2
    else (if v2 = lca then v1 else max v1 v2)
\end{ocaml}
\C{MList(Qty)}, like \C{MList(Int)}, is a list of integers, and has
the same semantics under a sequential execution. Their differences
however manifest under a concurrent execution, where both adopt
different methods to reconcile concurrent updates.

@@

{\bf Shopping List}. A non-trivial demonstration of composability of
mergeable types is a shopping list application. The application allows
its users to collaboratively build a grocery shopping list (listing
items in the decreasing order of their priority), by adding new items
with a specified quantity, deleting existing items, and increasing or
decreasing their quantity. An item (\C{Item.t}) is represented as a
tuple of its name (a string) and the quantity (A \C{Qty.t}). The merge
function for
\begin{wrapfigure}{l}{.5\textwidth}
  \centering
  \includegraphics[scale=0.3]{Figures/shopping-list}
  \caption{Merging concurrent shopping lists}
  \label{fig:shopping-list}
\end{wrapfigure}
\C{Item.t} merges the quantities of items (using \C{Qty.merge}) having
the same names. Items with different names are considered distinct. A
shopping list is a mergeable list of items (i.e.,
\C{MList(Item)}). \C{MList(Item).merge} automatically lifts the merge
semantics for items to merge semantics for lists of items.
Fig.~\ref{fig:shopping-list} illustrates its behavior.  Here, two
users, Alice and Bob, concurrently edit a shopping list whose initial
contents are \C{milk} and \C{eggs}. While both simultaneously update the
quantity of eggs, Bob also removes milk and inserts candy. The merged
shopping list contains eggs and candy, where the quantity of eggs is
obtained by merging Alice's and Bob's quantity as per the definition
of \C{Qty.merge}.

>>>>

\subsection{Ropes}

A rope is a data structure that faciliates efficient
implementations of operations such as concatenation, splitting, lookup
and insertion at a random position etc., for a list of values. The
values are usually characters; their lists being strings. The data
structure is a binary tree, whose leaf nodes store substrings of a
larger string such that their left-to-right (in-order) concatenation
yeilds the larger string. The idea is to facilitate efficient lookups
and insertions at random positions in the string by storing
position indexes in the internal nodes of the tree.

@@

\subsection{Lists}

@@

As an abstract data type, lists support three
operations: \C{insert x i l}, to insert an element \C{x} at position
\C{i} in the list \C{l}, \C{delete i l}, to delete the element at
\C{i}'th position, and \C{subst i x l}, to substitute the element at
the \C{i}'th position with \C{x}.  Lists of mergeable items are
mergeable if they preserve the following properties:
\begin{itemize}
  \item Elements present in both the merging lists will be present in
  the final list. No element is deleted unless it is deleted in at
  least one list.
  \item An element deleted in one list will not reappear in the merged
  list (unlike Amazon's replicated shopping cart~\cite{dynamo}).
  \item The order of elements is retained. That is, if $x$ occurs
  before $y$ in one of the lists, and if both elements are present in
  the merged list, then $x$ occurs before $y$ in the merged list.
\item If an element $y$ is substituted for an element $x$ in one list,
  then $y$ is also substituted for $x$ in the merged list (i.e., $x$
  is absent and $y$ is present), unless $x$ is deleted or substituted
  in the other list. In the former case, deletion wins, and neither
  $x$ nor $y$ occur in the merged list. In the latter case, if $x$ is
  substituted by a different element $z$ in the other list, then the
  merged list substitutes $x$ by a merge of $y$ and $z$, defined as
  per the merge semantics of the list element type.
\end{itemize}

@@

\subsection{Lists of Integers and Quantities}.

A mergeable list of integers can be composed using mergeable lists and
mergeable integers.  Concretely, it is a composition of \C{MList} and
\C{Int} types (i.e., \C{MList(Int)}), where \C{Int.merge} is the same
as \C{Counter.merge} (Sec.~\ref{sec:intro}).  An alternative
instantiation of \C{MList} is for a quantity type
\begin{wrapfigure}{l}{.45\textwidth}
  \centering
  \includegraphics[scale=0.4]{Figures/list-eg}
  \caption{Lists of mergeable values are mergeable. }
  \label{fig:list-eg}
\end{wrapfigure}
(\C{Qty}) that supports \C{add} and \C{subtract} operations, but
defines its \C{merge} opeation a little differently:
\begin{ocaml}
let merge lca v1 v2 = if v1 = lca then v2
    else (if v2 = lca then v1 else max v1 v2)
\end{ocaml}
\C{MList(Qty)}, like \C{MList(Int)}, is a list of integers, and has
the same semantics under a sequential execution. Their differences
however manifest under a concurrent execution, where both adopt
different methods to reconcile concurrent updates.

If efficient insertions and lookups (by position) are desired over a
large list of integers, while admitting concurrent updates, we require
a mergeable rope of integers, which can be obtained by simply
composing \C{MList(Int)} with \C{MRope} (i.e., \C{MRope(MList(Int))}).

@@

\subsection{Shopping List}

\begin{figure}
\centering
\begin{tabular}{l||l}
\begin{ocaml}
let alice_f : L.t unit t =
  fork bob_f >>= fun () -> (* Invite Bob *)
  get () >>= fun c0 ->
  let qty = L.assoc "eggs" in
  let c0' = L.subst 1 ("eggs",qty+1) c0 in
  sync () ~v:c0' >>=
  fun c1 -> return ()
\end{ocaml}
&
\begin{ocaml}
let bob_f : L.t unit t =
  get () >>= fun c0 ->
  let c0' = L.insert ("candy",Qty.of_int 1) 1 @@ 
            L.delete @@ L.lookup "eggs" in
  sync () ~v:c0' >>=
  fun c1 -> return ()
\end{ocaml}
\\
\end{tabular}
\caption{Bob and Alice collaboratively build a grocery shopping list}
\label{fig:shopping-list-code}
\end{figure}

A non-trivial demonstration of composability of mergeable types is a
shopping list application. The application allows its users to
collaboratively build a grocery shopping list (listing items in the
decreasing order of their priority), by adding new items with a
specified quantity, deleting existing items, and increasing or
decreasing their quantity. An item (\C{Item.t}) is represented as a
tuple of its name (a string) and the quantity (A \C{Qty.t}). The merge
function for
\begin{wrapfigure}{l}{.5\textwidth}
  \centering
  \includegraphics[scale=0.3]{Figures/shopping-list}
  \caption{Merging concurrent shopping lists}
  \label{fig:shopping-list}
\end{wrapfigure}
\C{Item.t} merges the quantities of items (using \C{Qty.merge}) having
the same names. Items with different names are considered distinct. A
shopping list is a mergeable list of items, i.e., \C{MList(Item)} (a
grocery shopping list may not be large enough to justify using ropes,
but \C{MRope(MList(Item))} is definitely a possibility).
\C{MList(Item).merge} automatically lifts the merge semantics for
items to merge semantics for lists of items.
Fig.~\ref{fig:shopping-list} illustrates its behavior.  Here, two
users, Alice and Bob, concurrently edit a shopping list whose initial
contents are \C{milk} and \C{eggs}. While both simultaneously update
the quantity of eggs, Bob also removes milk and inserts candy
(Fig.~\ref{fig:shopping-list-code} shows the \name code for the
concurrent sessions). The merged shopping list contains eggs and
candy, where the quantity of eggs is obtained by merging Alice's and
Bob's quantity as per the definition of \C{Qty.merge}.
